Convergence Rates for the MAP of an Exponential Family\\ and Stochastic Mirror Descent -- an Open Problem

### 2022-01-27 COLT2022/colt2022.tex
Add files for COLT. Many errors left

\documentclass[twoside]{article}

@@

%\pdfsuppresswarningpagegroup=1 %removes the warning
% This is a hack, the other way to remove is to convert all pdf into postscript or eps.
% RLP: not sure what is the point of this. With my latex it creates an error, and I do not have warnings, so I comment it out for now.

\usepackage[accepted]{aistats/aistats2022}
% If your paper is accepted, change the options for the package
% aistats2022 as follows:
%\usepackage[accepted]{aistats2022}

\setlength{\footskip}{3.31pt}

% If you set papersize explicitly, activate the following three lines:
%\special{papersize = 8.5in, 11in}
%\setlength{\pdfpageheight}{11in}
%\setlength{\pdfpagewidth}{8.5in}

%\let\oldsection\section
%\renewcommand{\section}[1]{\oldsection{\texorpdfstring{\uppercase{#1}}{#1}}}

% If you use natbib package, activate the following three lines:
\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}

%\usepackage{microtype} % RLP tried it, no change.
%\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\graphicspath{{figs/}}
%\usepackage{subfigure}
%\usepackage{subcaption}
%\usepackage{placeins}
\usepackage{hyperref}       % hyperlinks
\usepackage[dvipsnames]{xcolor}

@@

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newtheorem{problem}{Open Problem}

@@

@@

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
\runningtitle{Convergence Rates for the MAP of an Exponential Family and SMD -- an Open Problem}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
\runningauthor{R\'emi Le Priol, Frederik Kunstner, Damien Scieur, and Simon Lacoste-Julien}

\twocolumn[

\aistatstitle{Convergence Rates for the MAP of an Exponential Family\\ and Stochastic Mirror Descent -- an Open Problem}

\aistatsauthor{R\'emi Le Priol \And Frederik Kunstner}
\aistatsaddress{ 
	Mila, Université de Montreal 
	\And  University of British Columbia
	}

\aistatsauthor{Damien Scieur \And Simon Lacoste-Julien}
\aistatsaddress{Samsung, SAIT AI Lab, Montreal 
	\And Mila, Université de Montreal \\
	Samsung, SAIT AI Lab, Montreal\\
	Canada CIFAR AI Chair
}

]

@@

@@

\begin{figure}[t]
	\centering
\includegraphics[width=.4\textwidth]{gammaMAPrate.pdf}
	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
		Solid curve  are average over $10^5$ trials.
		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
		Shaded areas are 90\% confidence interval.
		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
	}
	\label{fig:curves}
\end{figure}

@@

\begin{figure*}[t]
	\centering
	\includegraphics[width=.8\textwidth]{figs/thales/numerical_schema_n=3.pdf}
	\caption{
	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MLE (blue) and MAP (orange) (\S\ref{ssec:gaussian} with $n=3$).
	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:bias-variance} from~\S\ref{ssec:bias-variance}.
	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the long edges of the triangle and the red curve.
	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution.
	}
	\label{fig:thales}
\end{figure*}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
	\caption{
	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. The asymptote is $\frac{1}{n}$.
	}
	\label{fig:gaussian_decomposition}
\end{figure}

@@

\subsubsection*{Acknowledgments}
We express our gratitude to Reza Babanezhad, Radu Dragomir and Hadrien Hendrikx for their insights on stochastic mirror descent. 
This work was partially supported by the NSERC Discovery Grant RGPIN2017-06936 and the Canada CIFAR AI Chair Program. 
Simon Lacoste-Julien is a CIFAR Associate Fellow in the Learning in Machines \& Brains program.

\bibliographystyle{apalike}
\bibliography{references}

\clearpage
\appendix
\onecolumn

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=\textwidth]{figs/optimal_n0.pdf}
	\caption{Contours of $\expect{\KL(\nat^*, \hat \nat_n)}$  for the Gaussian variance with $n\in\{1,5,10\}$, $\mu^*=1$ and $n_0, \mu_0$ spanning $[10^{-2}, 10^2]$. The expectation was estimated with $10^4$ draws for each value of $n_0, \mu_0$ We observe that the line $\frac{\mu_0}{\mu^*} = 1 + \inv{n_0}$ coincides with the bottom valley of this landscape.
	}
	\label{fig:optimal_n0}
\end{figure}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=.45\textwidth]{figs/2d/logpart.pdf}
	\includegraphics[width=.45\textwidth]{figs/2d/entropy.pdf}
	\caption{
	(left) Contours of the log-partition function~\eqref{eq:app-logpart}.
	(right) Contours of the entropy~\eqref{eq:app-entropy}.
	}
	\label{fig:logpart-entropy}
\end{figure}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.9\textwidth]{figs/2d/mirrormap.pdf}
	\caption{
	(top) Grid deformation produce by $\nabla \logpart(\nat)$.
	(bottom) Grid deformation produced by $\nabla\conj(\m)$.
	}
	\label{fig:mirrormaps}
\end{figure}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.9\textwidth]{figs/2d/gaussian_paths.pdf}
	\caption{
		Paths taken by MLE (blue) and MAP (orange) on top of contours for $\KL(\nat^* || \nat)$. 
		We set $\mu^*=(0, 1), \mu_0 = (1, 2)$, and $n_0=4$, and $n$ varies from $1$ to $20$.
		In green, red and purple, we represent the paths respectively taken by the MAP dual expectation  $(\bar\nat_n, \bar \mu_n)$, MAP primal expectation $(\tilde\nat_n,\tilde\mu_n)$, and MLE primal expectation. Recall that the MLE dual expectation is $\mu^*$ itself.
	}
	\label{fig:gaussian-paths}
\end{figure}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.4\textwidth]{figs/self_concordance/omega.eps}
	\includegraphics[width=0.4\textwidth]{figs/self_concordance/omega_star.eps}
	\caption{(left) Graph of the distance function $\omega$ and its dual $\omega^*$. (right) Graph of the dual of distance function $\omega^*$ evaluated at $\frac{t}{1-t}$, compared with $t^2$. The two curves cross each other at $t\approx 0.21$.}
	\label{fig:omega}
\end{figure}

>>>>

%\documentclass[anon,12pt]{colt2022} % Anonymized submission
\documentclass[final,12pt]{colt2022} % Include author names
% The following packages are automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e
\usepackage{times}

@@

%\usepackage{microtype} % RLP tried it, no change.
%\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\graphicspath{{../figs/}}
%\usepackage{subfigure}
%\usepackage{subcaption}
%\usepackage{placeins}
\usepackage{hyperref}       % hyperlinks
%\usepackage[dvipsnames]{xcolor}

@@

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{problem}{Open Problem}

@@

% Use \Name{Author Name} to specify the name.
% If the surname contains spaces, enclose the surname
% in braces, e.g. \Name{John {Smith Jones}} similarly
% if the name has a "von" part, e.g \Name{Jane {de Winter}}.
% If the first letter in the forenames is a diacritic
% enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

% Two authors with the same address
% \coltauthor{\Name{Author Name1} \Email{abc@sample.com}\and
%  \Name{Author Name2} \Email{xyz@sample.com}\\
%  \addr Address}

% Three or more authors with the same address:
% \coltauthor{\Name{Author Name1} \Email{an1@sample.com}\\
%  \Name{Author Name2} \Email{an2@sample.com}\\
%  \Name{Author Name3} \Email{an3@sample.com}\\
%  \addr Address}

% Authors with different addresses:
\coltauthor{%
 \Name{R\'emi {Le Priol}} \Email{remi.lp.17@gmail.com}\\
 \addr Mila, Université de Montréal
 \AND
 \Name{Damien Scieur} \Email{damien.scieur@gmail.com}\\
 \addr Samsung, SAIT AI Lab, Montreal
 \AND
 \Name{Frederik Kunstner} \Email{frederik.kunstner@gmail.com}\\
 \addr University of British Columbia
 \AND
 \Name{Simon Lacoste-Julien} \Email{slacoste@iro.umontreal.ca}\\
 \addr Mila, Université de Montreal \\
	Samsung, SAIT AI Lab, Montreal\\
	Canada CIFAR AI Chair
}
\title[%Short Title
	Convergence Rates for the MAP of an Exponential Family and SMD -- an Open Problem
	]{%Full Title of Article
	Convergence Rates for the MAP of an Exponential Family\\ and Stochastic Mirror Descent -- an Open Problem
	}

@@

\maketitle

@@

\begin{keywords}%
  List of keywords%
\end{keywords}

@@

\begin{figure}
	\centering
\includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
		Solid curve  are average over $10^5$ trials.
		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
		Shaded areas are 90\% confidence interval.
		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
	}
	\label{fig:curves}
\end{figure}

@@

\begin{figure*}[t]
	\centering
	\includegraphics[width=.8\textwidth]{thales/numerical_schema_n=3.pdf}
	\caption{
	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MLE (blue) and MAP (orange) (\S\ref{ssec:gaussian} with $n=3$).
	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:bias-variance} from~\S\ref{ssec:bias-variance}.
	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the long edges of the triangle and the red curve.
	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution.
	}
	\label{fig:thales}
\end{figure*}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=.4\textwidth]{gaussians/new_linear_n0=1.pdf}
	\caption{
	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. The asymptote is $\frac{1}{n}$.
	}
	\label{fig:gaussian_decomposition}
\end{figure}

@@

\acks{ % Acknowledgments---Will not appear in anonymized version
We express our gratitude to Reza Babanezhad, Radu Dragomir and Hadrien Hendrikx for their insights on stochastic mirror descent. 
This work was partially supported by the NSERC Discovery Grant RGPIN2017-06936 and the Canada CIFAR AI Chair Program. 
Simon Lacoste-Julien is a CIFAR Associate Fellow in the Learning in Machines \& Brains program.
}

\bibliographystyle{apalike}
\bibliography{../references}

\clearpage
\appendix

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=\textwidth]{optimal_n0.pdf}
	\caption{Contours of $\expect{\KL(\nat^*, \hat \nat_n)}$  for the Gaussian variance with $n\in\{1,5,10\}$, $\mu^*=1$ and $n_0, \mu_0$ spanning $[10^{-2}, 10^2]$. The expectation was estimated with $10^4$ draws for each value of $n_0, \mu_0$ We observe that the line $\frac{\mu_0}{\mu^*} = 1 + \inv{n_0}$ coincides with the bottom valley of this landscape.
	}
	\label{fig:optimal_n0}
\end{figure}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=.45\textwidth]{2d/logpart.pdf}
	\includegraphics[width=.45\textwidth]{2d/entropy.pdf}
	\caption{
	(left) Contours of the log-partition function~\eqref{eq:app-logpart}.
	(right) Contours of the entropy~\eqref{eq:app-entropy}.
	}
	\label{fig:logpart-entropy}
\end{figure}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.9\textwidth]{2d/mirrormap.pdf}
	\caption{
	(top) Grid deformation produce by $\nabla \logpart(\nat)$.
	(bottom) Grid deformation produced by $\nabla\conj(\m)$.
	}
	\label{fig:mirrormaps}
\end{figure}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.9\textwidth]{2d/gaussian_paths.pdf}
	\caption{
		Paths taken by MLE (blue) and MAP (orange) on top of contours for $\KL(\nat^* || \nat)$. 
		We set $\mu^*=(0, 1), \mu_0 = (1, 2)$, and $n_0=4$, and $n$ varies from $1$ to $20$.
		In green, red and purple, we represent the paths respectively taken by the MAP dual expectation  $(\bar\nat_n, \bar \mu_n)$, MAP primal expectation $(\tilde\nat_n,\tilde\mu_n)$, and MLE primal expectation. Recall that the MLE dual expectation is $\mu^*$ itself.
	}
	\label{fig:gaussian-paths}
\end{figure}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.4\textwidth]{self_concordance/omega.eps}
	\includegraphics[width=0.4\textwidth]{self_concordance/omega_star.eps}
	\caption{(left) Graph of the distance function $\omega$ and its dual $\omega^*$. (right) Graph of the dual of distance function $\omega^*$ evaluated at $\frac{t}{1-t}$, compared with $t^2$. The two curves cross each other at $t\approx 0.21$.}
	\label{fig:omega}
\end{figure}
